---
title: "Vis: Boxplots and Counts"
author: Zach del Rosario
date: 2020-06-04
output: github_document
dep: e-data02-derive
time: 30
reading: 0
---

# Vis: Boxplots and Counts

*Purpose*: *Boxplots* are a key tool for EDA. Like histograms, boxplots give us a sense of "shape" for a distribution. However, a boxplot is a *careful summary* of shape. This helps us pick out key features of a distribution, and enables easier comparison of different distributions.

*Reading*: (None, this is the reading)

```{r setup, include=FALSE}
# knitr options
knitr::opts_chunk$set(echo = TRUE)
```

```{r library}
library(tidyverse)
```

## Boxplots

Visuals like histograms, frequency polygons, and distributions give us a highly *detailed* view of our data. However, this can actually be overwhelming. To illustrate, let's dive straight into an exercise.

### **q1** Interpret plots

Which class of vehicle tends to have the most "middle" value of engine displacement (`displ`)? More importantly, which plot *best* helps you make that determination?

```{r q1-vis1}
## NOTE: No need to modify
# Density plot
mpg %>%
  ggplot(aes(displ, color = class)) +
  geom_density()
```

Note that the bold line in the middle of a boxplot is the *median* of the group.

```{r q1-vis2}
## NOTE: No need to modify
# Boxplot
mpg %>%
  mutate(class = fct_reorder(class, displ)) %>%
  # mutate(class = fct_reorder(class, displ, .fun = min)) %>%  #reorders around the min not med
  ggplot(aes(x = class, y = displ)) +
  geom_boxplot()
```

**Observations** - Which class tends to be most "middle" for `displ` values? - Which plot is more effective for answering the previous question?

-   The minivan tends to be the most middle of all of the cars

-   The bixplot has the best representation of this as it places the minivan class in the middle of all of the other boxplots/vehicles

## Boxplot definition

A boxplot shows a few key *summary statistics* from our data. The "box" itself shows the lower quartile (25% of the data) and upper quartile (75% of the data), while the bold line shows the median (50% of the data).

The following code shows how the quartiles can be manually computed.

```{r}
## NOTE: No need to edit
mpg %>% 
  filter(class == "midsize") %>% 
  
  ggplot(aes(x = class, y = displ)) +
  geom_hline(
    data = . %>%
      # Compute the quartiles
      summarize(
        displ_025 = quantile(displ, 0.25),
        displ_050 = quantile(displ, 0.50),
        displ_075 = quantile(displ, 0.75),
      ) %>%
      # Reshape the data for plotting
      pivot_longer(
        cols = contains("displ"),
        names_sep = "_",
        names_to = c(".value", "quantile")
      ),
    mapping = aes(yintercept = displ, color = quantile)
  ) +
  geom_boxplot()
```

The botplot also includes *fences* (the thin vertical lines) to show where there is some---but not very much---data. The boxplot also includes a heuristic for identifying outliers, which show up as dots.

## Reorganizing factors

There's a "trick" I've pulled in the earlier boxplot; I *reordered* the `class` variable based on the value of `displ` in each group. This is a way to make our plots more informative. The `fct_reorder(fct, x)` function is used in a `mutate()` call to directly override the original `fct` column.

### **q2** Reorder

Reorder `class` according to `hwy`. Pay attention to how this modifies the plot.

```{r q2-task}
## TASK: Edit this code to reorder `class` according to `hwy`
mpg %>%
  mutate(class = fct_reorder(class, hwy)) %>% 
  ggplot(aes(x = class, y = hwy)) +
  geom_boxplot()
```

## Cut helpers

Plotting multiple boxplots works best when we have a categorical variable for grouping. However, we can "hack" a continuous variable into a categorical one by "cutting" the values, much like when we bin values in a histogram. The following helpers give us different ways to cut a continuous variable:

-   `cut_interval()`
-   `cut_number()`
-   `cut_width()`

### **q3** Cut a continuous variable

Use a `cut_*` verb (of your choice) to create a categorical variable out of `carat`. Tweak the settings in your cut and document your observations.

*Hint*: Recall that we learned how to look up documentation in an earlier exercise!

```{r q1-task}
diamonds %>%
  ## TODO: Use mutate to add the column `carat_cut`
  mutate(carat_cut = cut_interval(carat, 10)) %>%  #large widths may not work for every data set
  ggplot(aes(x = carat_cut, y = price)) +
  geom_boxplot()
```

**Observations** - Write your observations here! - Make sure to distinguish between the trend in median (bar), quantiles (box edges), and whiskers (vertical lines)

-   price generally increases with the carat

-   The IQR is small for really small or really big carats (ex. 0.2-0.68 or 4.53 to 5.01)

-   There is a lot of price variability with the upper and lower quantiles the bigger the carat (especially within the 3 carat range)

-   The price max seems like it is realtively stable amongst most all carats

## Coordinate flipping

One last visual trick: Boxplots in ggplot are usually vertically oriented. However, we can *flip* the plot to give them a horizontal orientation. Let's look at an example:

```{r}
mpg %>%
  ggplot(aes(class, displ)) +
  geom_boxplot() +
  coord_flip()
```

Coordinate flipping is especially helpful when we have a lot of categories.

### **q4** Flip a plot

The following plot has overlapping x-axis labels. Use a verb from the reading to `flip` the coordinates and improve readability.

```{r q4-task}
mpg %>%
  ggplot(aes(manufacturer, hwy)) +
  geom_boxplot() +
  coord_flip() # good when x-axis names are long
```

This is a simple---but important---trick to remember when visualizing data with many categories.

<!-- include-exit-ticket -->

# Exit Ticket

<!-- -------------------------------------------------- -->

Once you have completed this exercise, make sure to fill out the **exit ticket survey**, [linked here](https://docs.google.com/forms/d/e/1FAIpQLSeuq2LFIwWcm05e8-JU84A3irdEL7JkXhMq5Xtoalib36LFHw/viewform?usp=pp_url&entry.693978880=e-vis03-boxplots-assignment.Rmd).
