---
title: "RMS Titanic"
author: "Katherine Danielson"
date: 02-02-2025
output:
  github_document:
    toc: true
prerequisites:
  - e-vis01-bar-charts
---

*Purpose*: Most datasets have at least a few variables. Part of our task in analyzing a dataset is to understand trends as they vary across these different variables. Unless we're careful and thorough, we can easily miss these patterns. In this challenge you'll analyze a dataset with a small number of categorical variables and try to find differences among the groups.

*Reading*: (Optional) [Wikipedia article](https://en.wikipedia.org/wiki/RMS_Titanic) on the RMS Titanic.

<!-- include-rubric -->

# Grading Rubric

<!-- -------------------------------------------------- -->

Unlike exercises, **challenges will be graded**. The following rubrics define how you will be graded, both on an individual and team basis.

## Individual

<!-- ------------------------- -->

| Category | Needs Improvement | Satisfactory |
|------------------|-----------------------------|-------------------------|
| Effort | Some task **q**'s left unattempted | All task **q**'s attempted |
| Observed | Did not document observations, or observations incorrect | Documented correct observations based on analysis |
| Supported | Some observations not clearly supported by analysis | All observations clearly supported by analysis (table, graph, etc.) |
| Assessed | Observations include claims not supported by the data, or reflect a level of certainty not warranted by the data | Observations are appropriately qualified by the quality & relevance of the data and (in)conclusiveness of the support |
| Specified | Uses the phrase "more data are necessary" without clarification | Any statement that "more data are necessary" specifies which *specific* data are needed to answer what *specific* question |
| Code Styled | Violations of the [style guide](https://style.tidyverse.org/) hinder readability | Code sufficiently close to the [style guide](https://style.tidyverse.org/) |

## Submission

<!-- ------------------------- -->

Make sure to commit both the challenge report (`report.md` file) and supporting files (`report_files/` folder) when you are done! Then submit a link to Canvas. **Your Challenge submission is not complete without all files uploaded to GitHub.**

```{r setup}
library(tidyverse)

df_titanic <- as_tibble(Titanic)
```

*Background*: The RMS Titanic sank on its maiden voyage in 1912; about 67% of its passengers died.

# First Look

<!-- -------------------------------------------------- -->

### **q1** Perform a glimpse of `df_titanic`. What variables are in this dataset?

```{r q1-task}
## TASK: Perform a `glimpse` of df_titanic
df_titanic %>% 
  glimpse()
```

**Observations**:

-   The main variables in the df_titanic() dataset are Class, Sex, Age, Survived, and "n" which denotes the number of people on the Titanic that fell under a certain classification. The "Class" variable is formed of four main facets: 1st class, 2nd class, 3rd class and crew. The "Sex" variable is a binary of male and female. The "Age" variable is broken down into two categories adult and child. The survived category simply notes a "yes" or "no" on whether or not people survived. When combining these different variables, we get a classification of a person based, and the "n" variable notes how many people that fit that exact classification survived. For example, there were no people who were a 1st class male child that did not survive.
-   There are 5 columns and 32 rows – suggesting 32 different classifications of individuals. Looking briefly, it appears a large portion of individuals died.

### **q2** Skim the [Wikipedia article](https://en.wikipedia.org/wiki/RMS_Titanic) on the RMS Titanic, and look for a total count of souls aboard. Compare against the total computed below. Are there any differences? Are those differences large or small? What might account for those differences?

```{r q2-summary}
## NOTE: No need to edit! We'll cover how to
## do this calculation in a later exercise.
df_titanic %>% summarize(total = sum(n))
```

**Observations**:

-   The Wikipedia article notes that there were an estimated 2,224 passengers and crew aboard. Out of this, they express approximately 1,500 individuals died. Comparatively, the Titanic data set only accounts for 2,201 individuals.
-   Are there any differences?
    -   There are differences between the Wikipedia article and df_titanic() data set. As the Wikipedia expresses there were 2,224 passengers and crew on board and the data set expresses only 2,201 people on board, there is a difference of 23 individuals. This means that there is a small difference of 23 people. Assuming the Wikipedia numbers are correct, the data set is missing only \~1% of its individuals.
-   If yes, what might account for those differences?
    -   As the data set has a variety of different variables, it is possible that for the 23 people not accounted for, certain or all variables may not have been able to be found as tracking during that time was limited. Other possibilities for the differences in the data set quantity of individuals may be due to faulty numbers on Wikipedia. As Wikipedia is not the most trusted source, these numbers may be slightly off and the data set could be pulling from a different location.

### **q3** Create a plot showing the count of persons who *did* survive, along with aesthetics for `Class` and `Sex`. Document your observations below.

*Note*: There are many ways to do this.

```{r q3-task}
## TASK: Visualize counts against `Class` and `Sex`
df_titanic %>% 
  filter (Survived == "Yes") %>%       #filter by survived
  ggplot(aes(Class, n, fill = Sex)) +  #assigns variables
  geom_col(position = "dodge")         #unstacks male and female data

#You do not need to group_by or sum(n) as when plotting with aes it will be filtering the categories and pulling anyone who fits the sex, class together -- inherently summing n
```

**Observations**:

-   2nd Class had the least number of survivors while the Crew had the greatest number of survivors. (The fact that the largest number of survivors was from the crew surprised me–I wonder if they knew the boat better and how to get off, if they were given priority or if they were the largest number of people on board. To determine this, we could look at the proportion of the crew to the rest of the occupants.) Out of all of the other classes, 1st Class having the highest level of survival makes sense to me as I feel they would have been given priority.

-   In 1st and 2nd Class, the majority of individuals who survived were female while in 3rd Class, the division between male and female survivors seemed fairly equal. The Crew had a much larger percentage of male survivors than females. It would be interesting to look into the sex makeup of the passengers and crew to see if maybe the majority of the crew was male, so the low female survival was not due to many dying, but rather many not actually functioning as crew. This could also be further analyzed by looking at the proportion of survival within sex to see if there was a social prioritization of females over males.

-   By diving deeper and looking at the proportions overall and seeing how many people were actually in each class as well would be interesting. While this graph is useful, it does not contextualize the data in the larger whole and leaves a lot to be wondered.

# Deeper Look

<!-- -------------------------------------------------- -->

Raw counts give us a sense of totals, but they are not as useful for understanding differences between groups. This is because the differences we see in counts could be due to either the relative size of the group OR differences in outcomes for those groups. To make comparisons between groups, we should also consider *proportions*.[1]

The following code computes proportions within each `Class, Sex, Age` group.

```{r compute-prop}
## NOTE: No need to edit! We'll cover how to
## do this calculation in a later exercise.
df_prop <-
  df_titanic %>%
  group_by(Class, Sex, Age) %>%
  mutate(
    Total = sum(n),
    Prop = n / Total
  ) %>%
  ungroup()
df_prop
```

### **q4** Replicate your visual from q3, but display `Prop` in place of `n`. Document your observations, and note any new/different observations you make in comparison with q3. Is there anything *fishy* in your plot?

```{r q4-task}
df_prop %>% 
  filter(Survived == "Yes") %>% 
  ggplot(aes(Class, Prop, fill = Sex)) +
  geom_col(position = "dodge", color = "black")  #the black shows the outline and shows how adult and children are still overlayed

```

**Observations**:

-   The proportions of female vs. male are independent of one another and are thus best represented when using position = "dodge" as they compare who did survive to who didn't survive within the sex and class.
-   From this, we can see that it looks like 100% of males and females in both 1st and 2nd class survived (However, as we will see later in question 5, this is not actually the case). In third class, less than 50% of females in this class survived, and roughly only 25% of males survived. In third class, there was a lower proportion of surviving individuals. Due to a lack of conceptualization in total numbers of individuals, you cannot compare the proportions between classes and know for sure if less females survived in third class than in crew (which had a much higher survival rate) as the total population of females in third class could be much larger than that in crew. This is much more visible in the prior graph where it shows the total number of individuals who survived because you can compare the survival between classes. However, the first graph does not actually show the proportion of individuals who survived. It would be ideal to combine these graphs to see what proportion of the original whole survived in relation to the original numbers.
-   Is there anything *fishy* going on in your plot?
    -   This graph truly highlights the ratio of male-to-female survival and a bit of class-to-class survival. When looking at both third class and crew, a much higher proportion of the women present in that class survived. This highlights the social and cultural dynamic of women and children first in emergencies. Further, you can see a bit of class prioritization as well. Those who were in higher classes (first and second) had a much higher proportion of survivors. This can not only be viewed in the sense of prioritization of class, but also of status and money: people with more money essentially could pay to survive. To understand this fully, more data is necessary. We could better understand this by including age in the graphs and seeing if that has an influence on proportions of survival and breaking the data down a bit further.

### **q5** Create a plot showing the group-proportion of occupants who *did* survive, along with aesthetics for `Class`, `Sex`, *and* `Age`. Document your observations below.

*Hint*: Don't forget that you can use `facet_grid` to help consider additional variables!

```{r q5-task}
df_prop %>% 
  filter(Survived == "Yes") %>% 
  ggplot(aes(Class, Prop, fill = Sex)) +
  geom_col(position = "dodge") +
  facet_grid(~Age)
```

**Observations**:

-   This graph is the most descriptive and useful graph as it breaks down the indivudals who survived the titanic into class, age and sex. An overarching trend of more females than males surviving in all classes is seen. When viewing this graph, it can be seen that in first and second class, all of the children survived – both male and female. Largely, this visualization of the data suggests that children had a higher proportion of survival. Another trend that is exhibited is the trend based on class. It appears as though the higher the class, or the more money you paid, the more likely you are to survive. One last thing to note is that there were no children in crew.
-   An important thing to note is that this graph creates a bit of confusion with the dataset and possible code computations. When looking at the Q4 graph, it makes it appear as though everyone in first and second class survived. However, this is problematic as when you look at the df_prop data set and the graph created in Q5, not all adults, especially male, in first and second class survived. This could possibly due to the fact that children in both of those classes had a 100% survival rate – these proportions, when combined with adults (who had very low survival rates), if not manipulated correctly could lead to skewing in the data set. (I wonder if it somehow "stacked" the proportions of age as well which led to skewing–this seems unlikely though as the proportions do not exceed 1.0)
-   If you saw something *fishy* in q4 above, use your new plot to explain the fishy-ness.
    -   My hypothesis about there being a social and cultural bias of prioritizing females and children and those with wealth was proven correct by looking at this graph. This graph shows that a much higher proportion of females survived than the proportion of males, and a much higher proportion of children survived than adults. As mentioned before this is likely because children and women are put first due to cultural norms. Further, as seen, especially within the adult's section of the graphs, class came into play when looking at the proportion of survival. A higher proportion of the men and women in first class survived than those in second and than those in third. This is likely because those of higher status and class are protected in society.

# Notes

<!-- -------------------------------------------------- -->

[1] This is basically the same idea as [Dimensional Analysis](https://en.wikipedia.org/wiki/Dimensional_analysis); computing proportions is akin to non-dimensionalizing a quantity.
