---
title: "The Islands, Part 1: Design"
author: "Katherine Danielson"
date: 03-26-2025
output:
  github_document:
    toc: true
prerequisites:
  - e-stat12-randomization
---

*Purpose*: So far in this class, I've handed you each dataset. But much of the important work of statistics happens *before* data collection. Issues with data collection generally can't be fixed by fancy analysis after the fact. Therefore, in this *two part* challenge, you will first *plan* and then *execute* a statistical project.

To do this, we'll make use of The Islands, an extremely detailed simulation of a virtual population, developed by researchers at the University of Queensland. This is a fascinating sandbox that helps illustrate the difficulties and complexities of collecting real data, while still being more tractable than collecting data from real humans.

This is part 1 of 2. In this part you will *plan* your statistical project, particularly your data collection. Sound data collection is called *statistical design*.

<!-- include-rubric -->

# Grading Rubric

<!-- -------------------------------------------------- -->

Unlike exercises, **challenges will be graded**. The following rubrics define how you will be graded, both on an individual and team basis.

## Individual

<!-- ------------------------- -->

| Category | Needs Improvement | Satisfactory |
|------------------------|------------------------|------------------------|
| Effort | Some task **q**'s left unattempted | All task **q**'s attempted |
| Observed | Did not document observations, or observations incorrect | Documented correct observations based on analysis |
| Supported | Some observations not clearly supported by analysis | All observations clearly supported by analysis (table, graph, etc.) |
| Assessed | Observations include claims not supported by the data, or reflect a level of certainty not warranted by the data | Observations are appropriately qualified by the quality & relevance of the data and (in)conclusiveness of the support |
| Specified | Uses the phrase "more data are necessary" without clarification | Any statement that "more data are necessary" specifies which *specific* data are needed to answer what *specific* question |
| Code Styled | Violations of the [style guide](https://style.tidyverse.org/) hinder readability | Code sufficiently close to the [style guide](https://style.tidyverse.org/) |

## Submission

<!-- ------------------------- -->

Make sure to commit both the challenge report (`report.md` file) and supporting files (`report_files/` folder) when you are done! Then submit a link to Canvas. **Your Challenge submission is not complete without all files uploaded to GitHub.**

# Setup

<!-- ----------------------------------------------------------------------- -->

```{r setup}
library(tidyverse)
library(rsample)

filename_random <- "./data/helvig-random.csv"
```

### **q1** Orientation

Log into [The Islands](https://islands.smp.uq.edu.au/index.php) and head to the [Visitor Center](https://islands.smp.uq.edu.au/visitors.php) on Providence island (to the east). Watch the first three videos on the `Guides` tab.

*Note:* There is no deliverable for this task. BUT, you'll need to be oriented to The Islands in order to do *any* of the tasks below.

The Islands is an *incredibly* detailed simulation of a real population. Islanders are born, die, and move around. The Islands was designed as a teaching tool to help students learn how to apply statistics to collecting data in the real world.

### **q2** Occupied homes

Find the total number of homes in Helvig, and count the number of unoccupied homes. Answer the questions below.

NB. When Zach counted the homes (on 2025-01-08), he found 20 unoccupied out of 536 total. That means the occupied percentage was about 96%. But you still need to count, because your numbers *will* be different!

**Observations**:

-   What is the total number of homes in Helvig?
    -   555 total homes in Helvig
-   What is the number of unoccupied homes? (*Hint*: This is not given anywhere. You will have to count them!)
    -   Only 1 home in Helvig is unoccupied
-   What percent of homes are *occupied*?
    -   99.8% of the homes are occupied
-   Are there any sources of *real* uncertainty in the percent occupied you calculated?
    -   There is one main source of real uncertainty in the percent occupied I calculated. The first is time lag in data collection. That is, occcupancy rates may change by the time I submit this assignment (or between data collection and analysis) as this is a constantly updating graph and people can move in and out.
-   Are there any sources of *erroneous* uncertainty in the percent occupied you calculated?
    -   The main source of erroneous uncertainty is human counting error. If I did not recognize a house to be unoccupied, it would introduce errors into the percentage. Similarly, in other instances, double-counting would impact the results as well (however, as only one house was deemed unoccupied, this cannot be the case).

Zach looked at the first 25 homes in Helvig and recorded the `age` and `name` of every person in those homes. These people are provided in `helvig-seq.csv`.

```{r}
## NOTE: Do not edit this
df_sample_seq <- read_csv("./data/helvig-seq.csv")
df_sample_seq 
```

You'll use this dataset as a starting point to figure out prevalent *last names* in Helvig.

### **q3** Find common names: Sequential sample

Complete the code below to find the prevalence of the most common *last name* in Helvig using the sample `df_sample_seq`. Answer the questions below.

Make sure to include in `df_q3` the counts as a column named `n`, and the prevalence (number of occurrence divided by total people) as a column `p`. Also, make sure to sort the data in *descending* order of prevalence.

*Hint*: You will have to use what you've learned about string handling to extract the *last names* only!

```{r q3-task}
## TASK: Compute the prevalence and sort
df_q3 <- 
  df_sample_seq %>% 
  mutate(last_name = str_extract(name, "\\w+$")) %>%  # Extract last name
  group_by(last_name) %>%                             # Group by last name
  summarise(n = n()) %>%                              # Count occurrences
  mutate(p = n / sum(n)) %>%                          # Calculate prevalence
  arrange(desc(p))                                    # Sort by prevalence

df_q3
```

Use the following to check your work.

```{r q3-tests}
## NOTE: No need to change this
## Check that data has `p` column and is in descending order
assertthat::assert_that(
              all(df_q3 %>%
                 mutate(d = p - lead(p)) %>% 
                 filter(!is.na(d)) %>% 
                 pull(d) >= 0
              )
            )
print("Very good!")
```

*Observations*

-   What last name is most prevalent in `df_sample_seq`?
    -   The most prevalent last name in `df_sample_seq` is Sorensen
-   Is this sample representative of *all* houses in Helvig? Why or why not?
    -   No, this sample is not representative of all houses in Helvig. Firstly, this is a very small sample size for having a total of 555 houses–25 is only 4.5% of the total number of houses. Secondly, by choosing the first 25 houses alone, selection bias and a lack of randomness are introduced. That is, it is likely that if there are multiple people in a house, they will share the same last name. Additionally, it is highly plausible that individuals live near each other could be relatives. Due to the idea of a "neighborhood" like living style based on familial and demographic similarities, choosing only the first 25 homes in a row is not representative of all houses in Helvig.

In the exercises, we talked about the importance of random sampling. In the previous challenge, we were able to *simulate* a random sample by running a few lines of code. But in the real world, we have to work harder to gather a random sample. We'll do this in two stages: *plan* then *collect*.

### **q4** Plan a random sample

Complete the code below to draw a sample of size `n=25`. Replace `n_houses` with the (current) total number in Helvig.

```{r}
## TASK: Set the parameters for this code block
set.seed(101)   # Set a seed for reproducibility

## Select a random sample of houses
n_houses <- 555 # Total number of houses
n_sample <- 25  # Desired sample size


df_numbers_random <- 
  tibble(
    house = sample(
        1:n_houses,     # All integers from 1 to n_houses
        n_sample,       # Size of our sample
        replace = FALSE # Sample *WITHOUT* replacement
      )
  ) %>% 
  # Arrange for our data collection convenience
  arrange(house)

# Pull the column so we can list just the house numbers
df_numbers_random %>% 
  pull(house)
```

Use the following code to check your results.

```{r}
## NOTE: No need to change this
assertthat::assert_that(
  all(dim(df_numbers_random) == c(25, 1))
)
```

### **q5** Collect the random sample

Gather the names of all people in the homes you listed in the previous exercise. Match the same columns as `df_sample_seq`; those are, `house`, `age`, `name`. Make sure to include `NA` rows for houses with no occupants. Save your data as a CSV with the filename provided in the variable `filename_random`. Answer the questions below.

```{r}
## NOTE: Do not edit
filename_random
```

Note that this points to the `data/` subdirectory in your `challenges` folder.

The following code will load your data.

```{r}
## NOTE: Do not edit
df_sample_random <- 
  read_csv(filename_random)
```

Use the following to check your work.

```{r q5-tests}
## NOTE: No need to change this
# Check that the dataset has the correct column names
assertthat::assert_that(setequal(
  df_sample_random %>% names(),
  df_sample_seq %>% names()
))

# Check that all of the house numbers in the dataset match those that were planned
numVsamp <- 
  anti_join(
    df_numbers_random,
    df_sample_random %>% distinct(house),
    by = "house"
  ) %>% 
  pull(house)
assertthat::assert_that(
  length(numVsamp) == 0,
  msg = str_c("You are missing the houses: ", numVsamp)
)

sampVnum <- 
  anti_join(
    df_sample_random %>% distinct(house),
    df_numbers_random,
    by = "house"
  ) %>% 
  pull(house)
assertthat::assert_that(
  length(sampVnum) == 0,
  msg = str_c("You have extra houses: ", sampVnum)
)

print("Great work!")
```

*Observations*

-   Which sample---sequential or random---is more *representative* of all homes Helvig? Why?
    -   Random sampling is more representative of all homes in Helvig.
    -   Random sampling ensures that every home and individual in Helvig has an equal chance of being included in the sample. This ensures that there is no one preference towards individuals of a certain location or demographic. Further, this increases the chance of capturing the full diversity of Helvig and actually representing the composition of all islanders rather than 25 who live directly next to one another.

### **q6** Find common names: Random sample

Run the code below to find the prevalence of the most common *last name* in Helvig using the sample `df_sample_random`. Answer the questions below.

```{r}
# NOTE: No need to edit; run and answer the questions below
df_sample_random %>% 
  mutate(last = str_extract(name, "\\w+$")) %>% 
  count(last) %>% 
  arrange(desc(n)) %>% 
  mutate(p = n / sum(n))
```

*Observations*

-   Did you find any highly prevalent names using `df_sample_random` that you *didn't* find in q3 (using `df_sample_seq`)? Write them here.
    -   Collins and Regan and Wilson were all names that were highly popular in `df_sample_random` but not seen in `df_sample_seq`.
-   Is there any reason that people with the same last name might tend to *live near each other*?
    -   It is likely that individuals with the same last name may tend to live near each other as they could be family. A lot of families like to have their relatives close by so that they can take care of each other and see each other more often, so it is likely that the higher prevalence of certain names in `df_sample_seq` is due to this.

You should have found some difference between the sequential and random samples. This is because we're only working with a *sample*---a limited number of observations from the entire population. We could go to every single house in Helvig and determine *exactly* how many people of each name there are (that's called a *census*), but the point of statistical inference is that we can make statements about a population using only a sample. What makes statistical inference powerful is that we can determine *how confident* we should be in our results, based on the size of our sample.

To do this, we'll use the bootstrap approach that you saw in the exercise sequence. We'll start by building a helper function.

### **q7** Write a helper function

Complete the code below to write a helper function. Your function will compute the proportion of people in a sample that have a user-specified last name.

```{r}
## TASK: Write a helper function that takes a dataframe with full names 
#  (provided in a `name` column), removes any invalid rows, and computes the
#  proportion of individuals with the user-specified `last` name (returned
#  in an `estimate` column).
name_prevalence <- function(df, last = "Collins") {
  df %>% 
    ## TODO: Finish this code
   
  
    filter(!is.na(name)) %>%    # Remove invalid rows
    mutate(last_name = str_extract(name, "\\w+$")) %>%  # Extract last name
    summarise(
      estimate = mean(last_name == last, na.rm = TRUE)  # Compute proportion
    ) %>%
    mutate(term = "prevalence") %>%                     # Add term column
    select(term, estimate)
} 
```

Use the following code to check your results.

```{r}
## NOTE: No need to change this
# Find the most prevalent name in the data
last_most <- 
  df_sample_random %>% 
  mutate(last = str_extract(name, "\\w+$")) %>% 
  count(last) %>% 
  arrange(desc(n)) %>% 
  slice(1) %>% 
  pull(last)

# Ensure correct columns
assertthat::assert_that(
  setequal(
    tibble(name = c("James")) %>% name_prevalence(., last = "James") %>% names(),
    c("term", "estimate")
  ),
  msg = "Your code should result a dataframe with just two columns: `term` and `estimate`"
)

# Ensure NA handling
assertthat::assert_that(
  !(tibble(name = c(NA_character_, "James")) %>% 
    name_prevalence(., last = "James") %>% 
    pull(estimate) %>% 
    is.na()),
  msg = "Ensure your code properly ignores NA's"
)

# Check for correctness
assertthat::assert_that(
  name_prevalence(df_sample_random, last = last_most) %>% pull(estimate) ==
    mean(str_detect(df_sample_random$name, last_most), na.rm = TRUE),
  msg = "Your code computed the wrong value"
)

print("Nice!")
```

### **q8** Construct a bootstrap confidence interval

Choose a prevalent name that you found in q6. Use bootstrap resampling with your helper function to construct a confidence interval for the prevalence of that name. Answer the questions below.

*Hint*: We learned how to do resampling-based inference in `e-stat09`.

```{r}
# TASK: Complete the code below to compute a bootstrap-based confidence interval
set.seed(101)

df_interval_bootstrap <- 
  df_sample_random %>% 
  bootstraps(., times = 1000) %>% 
  mutate(
    estimate = map(
      splits,
      function(split_df) {
## TODO: Finish this code, using the name_prevalence() helper you implemented
## HINT: Remember that you need to use analysis() when operating on split_df
     analysis(split_df) %>%                         # Extract data from split
          name_prevalence(., "Collins")                # Get prevalence estimate
      }
    )
  ) %>% 
  mutate(estimate = map(estimate, as_tibble)) %>%
  ## NOTE: No need to edit this line; this uses your bootstrap sample to compute
  # a confidence `int`erval using the percentile method
  int_pctl(., estimate)

df_interval_bootstrap 
```

**Observations**:

-   What is highest possible prevalence for your chosen name, based on the confidence interval you constructed?
    -   The highest possible prevalence of my chosen name "Collins" based on the CI I constructed is 18.64407%.
-   Note that we used the *random* sample with the bootstrap procedure in this task. Could we use the bootstrap to make a confidence interval using the sequential sample (`df_sample_seq`) that would be representative of all of Helvig? Why or why not?
    -   No
    -   Bootstrapping assumes the original sample is representative of the population. Unfortunately, as we have discussed above, df_sample_seq is not representative of the whole population of Helvig as the sequential sampling strategy introduced a clustering bias for certain last names based often on neighborhood or familial relation. So while we technically could apply a bootstrap, it would not be representative of the whole population, as the n = 1000 resamplings that will occur will be with data that is not representative–thus, creating no way that the resampled data could be fully representative of Helvig. This instead would magnify any initial bias and create an overly narrow, unreliable and imprecise outcome.

### **q9** Discover possible measurements

Click on a single islander (you can do this from one of the houses), and take a look at what info they can provide. Write down three pieces of info that you're particularly interested in; this will inform your quantity of interest for the final task.

*Hint*: The videos from the [Visitor Center](https://islands.smp.uq.edu.au/visitors.php) will be especially helpful for getting some ideas.

-   Occupation in relation to gender
    -   occupation
    -   gender
-   Wage/worth in relation to age
    -   wage/worth
    -   age
-   Reaction time vs. age
    -   reaction time with ruler test
    -   age

### **q10** Planning a study (TEAMWORK)

Challenge 10 will be a team assignment where you plan and execute a statistical study using The Islands. You should meet with your learning team for Challenge 10 to complete the steps below---though everyone is responsible for including the plan in their individual Challenge 08 submission.

#### **Population**

-   We will be studying only the farmers of Helvig as our population.

#### **Quantity of interest**

-   Our main quantity of interest is worth.

**Covariates**

-   There are several covariates:

    -   Age

    -   Gender

    -   Occupation

#### **Observation or experiment?**

The Islands allows you to ask islanders to complete tasks. If you just take measurements on your participants, then it's an observational study. But if you also introduce something that's meant to change the outcome of a measurement (e.g., drinking coffee before taking a test), that's called an experimental study. You need to decide whether your study is observational or experimental.

-   This study is a strictly observational study as we are just collecting the data from our participants.

#### **Question / Hypothesis**

-   How are farming occupation and islander worth related for the residents of Helvig?

-   **We believe that certain farming occupations increase islander individual worth.** We are also investigating gender and age in conjunction to isolate the effects of farming occupation and worth.

#### **Sampling plan**

-   What steps will you take to collect the data?

There are five farming occupations in Helvig: Dairy, Oats, Pigs, Poultry, and Sheep. We copied and pasted the list from the Bureau of Helvig in the respective order, and the names are ordered alphabetically based on last names. This list is then put into an Excel spreadsheet, where the first person is enumerated as 1, and so on.

We will then put this into a random number generator to select the individuals for our sample based on assigned numbers. After the generation of these 45 random numbers, we will collect the data for each assigned person–occupation, worth, gender, and age. To do so, we will go to the Helvig Bureau, go to “Registers,” find each individual’s name under their farming occupation, click on their name, and note their gender, age and worth. This data will then go into a spreadsheet that we will later turn into a csv file for data analysis. 

As this is an observational study, not an experimental one, the introduction of bias is very limited as we are simply collecting four pieces of data from each person.

-   How will you ensure the data is representative of your chosen population?

We will take a random sample of 45 individuals (roughly half of the population). We placed all of the farming occupations and their names into a sheet. Once we do this, we will use a random number generator to select 45 individuals from the total farming population. By using a larger random sample, this will ensure we are able to capture demographics and information from all of the different farming groups (which range in size 10-31 individuals). This will additionally capture the variance in gender and age based on population demographics. 

-   How will you choose your sample size?

We will choose a sample size of 45 individuals as we want to be fairly confident in our results. As the population size is 92 individuals and we have four members on our team, each collecting roughly 11 people’s worth of data is attainable.
