---
title: "The Islands, Part 2: Study"
author: "Katherine Danielson, Olivia Smith, Dakota Chang, Cooper Penkava"
date: 04-10-2025
output:
  github_document:
    toc: true
prerequisites:
  - c10-islands-pt1
---

*Purpose*: This is part 2 of 2. In part 1 you *planed* your statistical project, particularly your data collection. In this part you will give updates on your plan, and report your findings.

This challenge is deliberately shorter so you have time to collect and analyze your data.

*Important note*: While we expect that you did your data collection with your team, you need to complete your own individual report for c10.

<!-- include-rubric -->

# Grading Rubric

<!-- -------------------------------------------------- -->

Unlike exercises, **challenges will be graded**. The following rubrics define how you will be graded, both on an individual and team basis.

## Individual

<!-- ------------------------- -->

| Category | Needs Improvement | Satisfactory |
|------------------------|------------------------|------------------------|
| Effort | Some task **q**'s left unattempted | All task **q**'s attempted |
| Observed | Did not document observations, or observations incorrect | Documented correct observations based on analysis |
| Supported | Some observations not clearly supported by analysis | All observations clearly supported by analysis (table, graph, etc.) |
| Assessed | Observations include claims not supported by the data, or reflect a level of certainty not warranted by the data | Observations are appropriately qualified by the quality & relevance of the data and (in)conclusiveness of the support |
| Specified | Uses the phrase "more data are necessary" without clarification | Any statement that "more data are necessary" specifies which *specific* data are needed to answer what *specific* question |
| Code Styled | Violations of the [style guide](https://style.tidyverse.org/) hinder readability | Code sufficiently close to the [style guide](https://style.tidyverse.org/) |

## Submission

<!-- ------------------------- -->

Make sure to commit both the challenge report (`report.md` file) and supporting files (`report_files/` folder) when you are done! Then submit a link to Canvas. **Your Challenge submission is not complete without all files uploaded to GitHub.**

# Setup

<!-- ----------------------------------------------------------------------- -->

```{r setup}
library(tidyverse)
library(rsample)
library(ggforce)

# TODO: Include any other packages you need
```

### **q1** Planning a study (TEAMWORK)

While you provided this plan in c08 (Part 1), please include your plan here. In particular, describe how you updated your plan in response to feedback.

#### **Population**

-   We will be studying only the farmers of Helvig as our population.

*We updated this to be uniform between all members–we want our population to be solely the farmers of Helvig.*

#### **Quantity of interest**

-   Our main quantity of interest is worth. This value can be viewed when clicking on an islander’s profile.

**Covariates**

-   There are several covariates. They can all be viewed by clicking on an islander’s profile:

    -   Age

    -   Gender

    -   Occupation

#### **Observation or experiment?**

The Islands allows you to ask islanders to complete tasks. If you just take measurements on your participants, then it's an observational study. But if you also introduce something that's meant to change the outcome of a measurement (e.g., drinking coffee before taking a test), that's called an experimental study. You need to decide whether your study is observational or experimental.

-   This study is a strictly observational study as we are just collecting the data from our participants.

#### **Question / Hypothesis**

-   How are farming occupation and islander worth related for the residents of Helvig?

-   **We believe that certain farming occupations increase islander individual worth.** We are also investigating gender and age in conjunction to isolate the effects of farming occupation and worth.

#### **Sampling plan**

-   What steps will you take to collect the data?

There are five farming occupations in Helvig: Dairy, Oats, Pigs, Poultry, and Sheep. We copied and pasted the list from the Bureau of Helvig in the respective order, and the names are ordered alphabetically based on last names. This list is then put into an Excel spreadsheet, where each person is located along with that islanders corresponding farming occupation.

We will then use R’s random number generator with a seed of 101 to select the individuals for our sample based on assigned numbers. We are going to choose 44 random islanders, stratified by the proportions of the population that each type of farming makes up within the whole of Helvig farmers. After the generation of these 44 stratified random numbers, we will collect the data for each assigned person–occupation, worth, gender, and age. To do so, we will go to the Helvig Bureau, go to “Registers,” find each individual’s name under their farming occupation, click on their name, and note their gender, age and worth. This data will then go into a spreadsheet that we will later turn into a csv file for data analysis. 

As this is an observational study, not an experimental one, the introduction of bias is very limited as we are simply collecting four pieces of data from each person.

-   How will you ensure the data is representative of your chosen population?

We will take a stratified random sample of 44 individuals (roughly half of the population). We placed all of the farming occupations and their names into a sheet. Once we do this, we randomly select 44 individuals from the (stratified) total farming population. By using a larger random sample, this will ensure we are able to capture demographics and information from all of the different farming groups (which range in size 10-31 individuals). This will additionally capture the variance in gender and age based on population demographics. 

-   How will you choose your sample size?

We will choose a sample size of 44 individuals as we want to be fairly confident in our results. As the population size is 92 individuals and we have four members on our team, each collecting roughly 11 people’s worth of data is attainable.

-   **How did the plan change?**

*We went away from the random number generator idea so that we could do a stratified random sample instead. For this, we just calculated the proportion that each type of farming makes of the whole of farming and then randomly selected those filtered populations up to that proportion times 44, our total sample size. We also specified how you can find the data in relation to the islander at hand.*

### **q2** EDA

Conduct an EDA on your data. Add as many code chunks as you need. Ensure your EDA gives enough context of the data for us to understand what comes next in this report.

**General Population EDA**

The general population EDA was performed to understand the quantities of each type of farmer and determine the proportions for our stratified sample.

```{r}
farmerdata_pop <- "./data/c10_farmerdata.csv"

#Load the general population data and rename
df_pop <-
  read_csv(
    farmerdata_pop,
  )

glimpse(df_pop)

#Identify the number of distinct names in the Full_Names
df_pop %>% 
  distinct(Name)

#Identify the different Farmer_Types and count for each of them 
df_pop %>%
  count(Type)
```

*Observations*

-   `Name`: This identifies the name of the farmer. There are a total of 92 farmers on Helvig. None of them hold two different kinds of farming jobs at once–they all have only one kind of farming that they do.

-   `Type`: This identifies the type of farmer. There are 5 different types of farmers: Dairy, Oat, Pig, Poultry, and Sheep.

    -   There are 10 Dairy farmers, 11 Sheep farmers, 15 Poultry farmers, 25 Pig farmers, and 31 Oat farmers.

-   We used the data collected here to create a proportional stratified sample. This stratified sample has an EDA performed as well below.

**Collection of Stratified Sampling**

Determining proportions of each type of farmer.

```{r}
# summarise df and count values of column type then mutate to include proportions 
df_summary <- df_pop %>%
  group_by(Type) %>%
  summarise(count = n()) %>%
  mutate(Prop = count / sum(count))

df_summary
```

Creation of the stratified sample and selection of the 44 people.

```{r}
# randomly draw numbers from each proportion count to get a stratified random sample

# set sample size
total_sample_size <- 45  

# calculate sample size per stratum
df_summary <- df_summary %>%
  mutate(sample_size = round(Prop * total_sample_size))

# stratified sampling
set.seed(101)

stratified_sample <- df_pop %>%
  group_by(Type) %>%
  group_split() %>%
  map2_df(df_summary$sample_size, ~ .x[sample(nrow(.x), .y), ])

# view result
stratified_sample

# export result
write_csv(stratified_sample, "stratified_sample.csv")

```

**Sample Population EDA**

This EDA is on the stratified sample.

```{r}
farmerdata_samp <- "./data/c10_farmersample.csv"

#Load the general population data and rename
df_samp <-
  read_csv(
    farmerdata_samp,
  ) %>% 
  rename(
    Worth = "Net Worth"
  )

glimpse(df_samp)

#Idenitfy the number of farmers in each Type
df_samp %>% 
  count(Type)

#Idenitfy the number of each Gender
df_samp %>% 
  count(Gender)

#Identify the mean, sd and IQR of Age
df_samp %>% 
  summarize(
    min_age = min(Age),
    max_age = max(Age),
    mean_age = mean(Age),
    sd_age = sd(Age),
    IQR_age = IQR(Age)
  )

#Identify the mean, sd and IQR of Worth
df_samp %>% 
  summarize(
    min_worth = min(Worth),
    max_worth = max(Worth),
    mean_worth = mean(Worth),
    sd_worth = sd(Worth),
    IQR_worth = IQR(Worth)
  )
```

*Observations*

-   `Name`: There are names (first and last name) of the 44 indiviudals in the sample. They represent farmers from five different `Type`s of farming on Helvig.

-   `Type`: This represents the five different types of farming in Helvig: Dairy farming, Oat farming, Pig farming, Poultry farming and Sheep farming. The number of individuals from each farming `Type` was selected using a proportional stratified sample.

    -   Dairy farming has 5 individuals represented in the sample

    -   Oat farming has 15 individuals represented in the sample

    -   Pig farming has 12 individuals represented in the sample

    -   Poultry farming has 7 individuals represented in the sample

    -   Sheep farming has 5 individuals represented in the sample

-   `Gender`: This represents the gender of the individual and is (from what we sample) a gender binary

    -   There are 18 females

    -   There are 26 males

-   `Net Worth` / `Worth`: This variable was changed to `Worth` and represents the net worth of individuals in the sample. This is a widely ranging number with the maximum being \~12.6 times higher than the minimum.

    -   The minimum worth is \$794

    -   The maximum worth is \$10034

    -   The sample has a mean worth of \$3734.864 and has a high standard deviation of \$2391.152 with an IQR of \$3073.25.

-   `Age`: This variable illustrates the age of each individual in the sample, and ranges quite widely.

    -   The minimum age is 18

    -   The maximum age is 59

    -   The mean age is 31.22727 with a standard deviation of 10.09689 and an IQR of 14.25.

### **q3** Key Analyses

Present the key analyses that support your questions / hypotheses. This could include summary statistics (e.g., a proportion of Islanders), grouped summary statistics (e.g., a proportion for group A, and for group B), or visualizations (e.g., a histogram). This section should be short, and every analysis should directly relate to q4.

```{r}
df_samp %>% 
  mutate(Type = fct_reorder(Type, Worth)) %>% 
  ggplot(aes(Type, Worth)) +
  geom_boxplot() +
  geom_point(color = "blue", alpha = 0.2) +
  labs(
    x = "Farming Type",
    y = "Farmer Net Worth",
    title = "Net Worth of Different Types of Farmers on Helvig"
  ) +
  theme_minimal()

```

```{r}
anova_result_WT <- aov(Worth ~ Type, data = df_samp)
summary(anova_result_WT)
```

*Observations*

-   As illustrated by the anova test there is no statistically significant difference in `Worth` between farmer `Type`. The F-value is 0.169 showing very little difference between group means compared to the variation within groups. The p-value of 0.953 shows there is no statistical significance between `Type` and `Worth`.
-   While there is not a statistically significant difference between farming-type median net `Worth`s, there is a trend in the median earnings. Oat farming generally has the lowest median `Worth` – coming in below 2500. However, while Oat farming produces the lowest median new `Worth`, it also contains the highest outlier. Dairy, Sheep, and Pig farming all have very similar median `Worth` values, however, Dairy makes the lowest median of the three, then Sheep, then Pig. Poultry has a decently higher median `Worth`, but has a smaller sample size compared to that of Oat and Pig.
-   All types of farming have quite a large distribution of individual worth. Specifically, the largest distribution of `Worth` appears to be within Oat and Pig farming. It should be noted that these two Types of farming had the largest proportional stratified sample size at 15 and 12, respectively. The remaining three have smaller distributions, with Poultry having the smallest distribution. Poultry had a smaller sample size (7), but has the highest median net Worth and has 5 very concentrated points. It would be interesting to look at the total population to see if this trend remains the same.
    -   One other interesting thing of note in this graph is the concentration of individuals within Oat and Pig farming (the two largest distributions). Both of them appear to have a lower and upper cluster within the box plot. It would be interesting to see what possibly causes this clustering–whether it be age, gender, etc..

```{r}
# Boxplot of Worth by Gender
worth_boxplot <- ggplot(df_samp, aes(x = Gender, y = Worth, fill = Gender)) +
  geom_boxplot() +
  labs(title = "Net Worth Distribution by Gender",
       x = "Gender",
       y = "Net Worth") +
  scale_fill_manual(values = c("F" = "pink", "M" = "lightblue")) +
  theme_minimal()

# Count plot of farmer Type by Gender
type_count_plot <- ggplot(df_samp, aes(x = Type, fill = Gender)) +
  geom_bar(position = "dodge") +
  labs(title = "Count of Farmer Types by Gender",
       x = "Farmer Type",
       y = "Count") +
  scale_fill_manual(values = c("F" = "pink", "M" = "lightblue")) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))


df_perc <- df_samp %>%
  count(Type, Gender) %>%  
  group_by(Gender) %>%         
  mutate(Percentage = n / sum(n) * 100) %>%
  ungroup()                       

type_percentage_plot <- ggplot(df_perc, aes(x = Type, y = Percentage, fill = Gender)) +
  geom_col(position = "dodge") +  # Dodged bars (side-by-side)
  geom_text(
    aes(label = paste0(round(Percentage, 1), "%")),  # Add % labels
    position = position_dodge(width = 0.9),          # Align with bars
    vjust = -0.5,                                    # Place above bars
    size = 3
  ) +
  scale_y_continuous(
    limits = c(0, 45), 
    expand = expansion(mult = c(0, 0.1))  
  ) +
  labs(
    title = "Percentage of Each Gender by Farmer Type",
    x = "Farmer Type",
    y = "Percentage (%)"
  ) +
  scale_fill_manual(values = c("F" = "pink", "M" = "lightblue")) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

anova_gender <-
  aov(Worth ~ Gender + Type, df_samp)

anova_type <- 
  aov(
    Gender_bool ~ Type, 
    df_samp %>% 
      mutate(Gender_bool = (Gender == "F"))
  )

summary(anova_gender)
summary(anova_type)


print(worth_boxplot)
print(type_count_plot)
print(type_percentage_plot)
```

*Observations*

-   The box plot shows a slight difference between the median and bounds on the IQR between male and female islanders, although it is quite slight. Given the fact that there are only 22 data points here, it is hard to tell if this difference is significant or not. According to the anova we ran, there is a p-value of 0.413, which indicates that this is not significant enough to raise any flags. So, the data collected by us gives the impression that there is not significant enough data difference to come to the conclusion that there is a difference between worth across gender.

-   The dodged bar chart shows that there is also not a very significant difference in the counts of the two - just off by one for most categories except pig. But we realized that plotting counts might not be absolutely the most real way of representing the data we collected - because of random sampling, we may not have gotten the same amount of males as females, so we decided to create the percentages plot.

-   After making it based on percentages, some differences become clearer - because of the differing counts, the percentages for oat and pig farmers are a little different actually. That wasn’t so clear before with just the counts! Running another anova shows that there is not really a significant correlation between male and female and the type of farmer (P = 0.977), so we just keep going.

```{r}
# age distribution of farmers in Helvig
df_samp %>% 
  ggplot(aes(Age)) +
  geom_histogram(bins = 10, fill = "#aaccff", color = "black") +
  labs(
    x = "Age",
    y = "Count",
    title = "Age of Farmers in Helvig"
  ) +
  theme_minimal()

# age distribution over different types
df_samp %>% 
  mutate(Type = fct_reorder(Type, Age)) %>% 
  ggplot(aes(Type, Age)) +
  geom_boxplot() +
  geom_point(color = "#aaccff") +
  labs(
    x = "Farming Type",
    y = "Age",
    title = "Age of Different Types of Farmers in Helvig"
  ) +
  theme_minimal()

# age distribution over different types sina plot

df_samp %>% 
  mutate(Type = fct_reorder(Type, Age)) %>% 
  ggplot(aes(Type, Age)) +
  geom_violin() +
  geom_sina(color = "#aaccff") +
  labs(
    x = "Farming Type",
    y = "Age",
    title = "Age of Different Types of Farmers in Helvig"
  ) +
  theme_minimal()
  
```

*Observations*

-   Farmers in Helvig tend to lean on the younger side, with most of them being between 18-38, with a spike at 48. There are no farmers above the age of 59.

-   There is little to no correlation between age and type of farming, this is found in both the box plot and also the anova analysis, where the p-value is high at 0.949. This indicates that there is no statistically significant difference between the ages of the different types of farmers.

-   In the sina plot, we can see that there is a slight clustering of ages within the Sheep farmers. However, that might be attributed to the small sample size of 5 or possible age distribution as noted in the prior graph.

```{r}
m <-
  df_samp %>% 
  filter(Gender == "M") %>% 
  tibble(
    age_bucket = cut_width(Age, width = 5, boundary = .0)
  ) %>% 
  group_by(age_bucket) %>% 
  summarize(n = -n()) %>% 
  mutate(Gender = "m")

f <- 
  df_samp %>% 
  filter(Gender == "F") %>% 
  tibble(
   age_bucket = cut_width(Age, width = 5, boundary = .0)
  )  %>% 
  group_by(age_bucket) %>% 
  summarise(n = n()) %>% 
  mutate(Gender = "f")


bind_rows(m, f) %>% 
  ggplot() +
  geom_col(aes(x = age_bucket, y = n, fill = Gender), position = 'stack',
                 color = 'black') +
  scale_y_continuous(
    breaks = c(-8, -4, -1, +1, +4, +8),
    labels = c("8", "4", "1", "1", "4", "8")
  ) +
  labs(
    title = "Gender Distribution Over Age",
    y = "Number of Farmers",
    x = "Age Bucket"
  ) +
  coord_flip()
  
  

g <-
  df_samp %>% 
  ggplot(aes(x = Age, y = Worth)) +
  geom_point(
    aes(x = Age, y = Worth, color = Type)
  ) + 
  geom_smooth(method = "lm") +
  labs(title = "Worth vs. Age") 

g + facet_grid(rows = vars(Gender))


```

*Observations*

Gender Distribution Over Age

-   The ratio between female and male farmers is relatively even, except for the 25 to 30 age range where there are seven times more men than women. 

-   The most farmers are in the 20-30 age range

-   There is only one farmer in the 50-60 age range

-   Female farmers have the highest probability of being in the 20-25 age range

-   Male farmers have the highest probability of being in the 25-30 age range

Worth vs. Age

-   Men and women farmers have roughly the same progression of net worth

-   Male farmers have a slightly lower standard error in their predicted wealth than female farmers

-   Farmers’ wealth does not seem to correlate with farming type or gender

-   Predicted worth increases with age, suggesting farmers can accumulate wealth throughout their life

-   For both male and female farmers, the max worth is a data point around \$10,000. Perhaps this is the max amount of wealth one can accumulate in Helvig. 

### **q4** Answers

Connect your Key Analyses to your questions / hypotheses. What did you conclude from your study?

Our guiding question for this analysis was: "How are farming occupation and islander worth related for the residents of Helvig?" In turn, this led to a null hypothesis of:

-   There is no statistical difference between the `Worth` in various farming `Type`s in Helvig.

From our key analyses about we fail to reject the null hypothesis.

As seen from our graph "Net Worth of Different Types of Farmers on Helvig," and its consequent ANOVA, the p-value is 0.953, meaning that there is no statistical significance between farming `Type` and `Worth`. While there is a small trend in the median as illustrated in the box plot, no true statistical trend can be captured. Thus, we decided to analyze other factors including `Gender` and `Age` to see their impact on each other and farming `Type` and `Worth`.

When looking at gender, we also can fail to show any real correlation between it and occupation or worth, as seen with the plots “Net Worth Distribution by Gender” and “Percentage of Each Gender by Farmer Type,” along with their associated ANOVA tests (p-values of .433, .957, and .977). The plots also visually describe a lack of correlation besides vague and minor differences, most likely do to random sampling.

In terms of `Age`, we see virtually no difference across different `Type`s of farmers. However, we do see that the farmers of Helvig lean towards the younger side, with most being between the ages of 18-38.

There is not a significant discrepancy between `Gender` proportion in the wider farming profession too. In the `Gender` Distribution Over `Age` histogram, we can see that there are roughly the same proportions of men and women farmers. There are a few outlier `Age` ranges, for example, the 25-30 age range where there are 7 times as many men as women, suggesting some sort of trend in farming popularity amongst genders. However, taken as a whole there are roughly the same amount of men and women in farming. This further suggests that gender is not a factor in farmer demographic. However, from the graph we can see that most farmers are clustered near the bottom in the younger age ranges, again suggesting farmers tend to be younger in age. 

Further, from the `Worth` vs. `Age` graph, we can see that men and women in every farming distribution can accumulate wealth with roughly the same effectiveness. The trend line has a similar slope for both genders, and the farming occupations do not seem to be correlated with worth or gender in both graphs. This further suggests that `Worth` is not affected by `Gender` or `Type`, but is instead by the accumulation of wealth over time, i.e. age. 

Despite all this analysis, `Worth` appears to be largely unaffected by other variables on Helvig and there appears to be very little covariance other than between the variables of `Worth` and `Age`. We continue to fail to reject our null hypothesis and conclude that different farming `Type`s on Helvig do not have an impact on individual `Worth`.
